dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format

### 2023-12-15 ms.Qmd
Update MS based on input from WI

Taxa fundamentally represent scientific hypotheses [@Sluys2013].
As such, taxonomic data are continuously being updated due to, for example, discovery of new species or revision of existing taxonomic concepts.
These taxonomic changes can be complicated.
For example, changing the status of one name to the synonym of another could potentially involve modifying many other names if they are synonyms of the first.
Because of these properties, maintaining a taxonomic database in compliance with DwC can be onerous and error prone if done by hand, especially if the database is large.
There is therefore a need for software to automate entry and validation of DwC taxonomic data.

@@

`dct_validate()` carries out automated data validation to ensure that the taxonomic database is in compliance with DwC.
At a minimum, this involves checking that the names of columns (terms) match those used by DwC.
While DwC makes recommendations for values of some terms, for most of these it does not specify the actual values allowed.
However, it is typical for taxonomic databases to make assumptions about the data based on their values.
For example, we would expect that every synonym should be linked to to an accepted name.
Therefore, validation in `dwctaxon` is flexible, allowing the user to select from multiple checks.

@@

- `check_mapping_strict`: Check that mapping of names follow expectations based on their taxonomic status (e.g., synonyms must map to accepted names, accepted names cannot have an `acceptedNameUsageID`, etc.).

@@

The are multiple options for the output of `dct_validate()`:

@@

Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
This research supported in part by Japan Society for the Promotion of Science (Kakenhi) grant no. 16H06279, 22H04925 and 22K15171.
Thanks to Collin Schwantes and Stephen Formel for reviewing the code and Noam Ross for handling the code review process at ROpenSci.

>>>>

Taxonomic data are frequently being updated due to, for example, discovery of new species or revision of existing taxonomic concepts.
These taxonomic changes can be complicated.
For example, changing the status of one name to the synonym of another could potentially involve modifying many other names if they are synonyms of the first.
Because of these properties, maintaining a taxonomic database in compliance with DwC can be onerous and error-prone if done by hand, especially if the database is large.
There is therefore a need for software to automate entry and validation of DwC taxonomic data.

@@

`dct_validate()` carries out automated data validation to ensure that the taxonomic database is in compliance with DwC.
At a minimum, this involves checking that the names of columns (terms) match those used by DwC.
While DwC makes recommendations for values of some terms, for most of these it does not specify the actual values allowed.
However, it is typical for taxonomic databases to make assumptions about the data based on their values.
For example, we would expect that every synonym should be linked to an accepted name.
Therefore, validation in `dwctaxon` is flexible, allowing the user to select from multiple checks.

@@

- `check_mapping_strict`: Check that mapping of names follows expectations based on their taxonomic status (e.g., synonyms must map to accepted names, accepted names cannot have an `acceptedNameUsageID`, etc.).

@@

There are multiple options for the output of `dct_validate()`:

@@

Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
This was research supported in part by Japan Society for the Promotion of Science (Kakenhi) grant no. 16H06279, 22H04925 and 22K15171.
Thanks to Collin Schwantes and Stephen Formel for reviewing the code and Noam Ross for handling the code review process at ROpenSci.
